Fail Faster

### 2025-03-16 paper/paper.tex
wibbles

Property-based testing (PBT) is a technique for software testing which uses
random test inputs to evaluate logical properties that the system is expected to
satisfy.  Recent studies on PBT usage~\cite{inpractice} have shown that
practitioners interleave PBT with their software development process, sometimes
testing their properties every time they save their code. This workflow makes
sense because the developer who just wrote the code is often the best person to
fix a potental bug; if a bug is found later, it will need to be fixed by someone
with less context.  All of this means that PBT is often run with very short time
budgets --- e.g., between 50 milliseconds and 30 seconds --- and it is important
that bugs can be found in that time-frame.
\jwc{phrasing could suggest ``race against time'' aspect.}\hg{Agree, can't see
how to fit this in yet}

@@

We find that the performance of existing generator libraries falls short of what
is possible. This is due to two major sources of inefficiency.
First, the same high-level design that gives generator libraries their
flexibility and expressiveness also introduces overhead: layered abstractions
and indirections hinder the compiler's optimizations, shifting more work to
runtime.
Second, the \rand that underlies the generator's operation --- the source of
random numbers that are used to make choices in the generator --- 
can constitute an unexpectedly high proportion of a generator's runtime.
\iflater
\hg{TODO later, I'm not 100\% sure it's clear how this particular question is
raised by the above. We could strengthen that connection}
\fi
These observations together raise a fundamental question: can generator
libraries produce code that is both expressive \textit{and} efficient?
%In practice, PBT is used 
%as a lightweight, ad hoc approach to software correctness,
%\jwc{Move this bit about how it's used to the first sentence. I.e. PBT is a testing technqique where random data is fed to a system under test...}
%aimed at 
%detecting bugs within a short testing window---typically 50 milliseconds 
%to 30 seconds. \jwc{This sells us a little short: we should say something to the effect of ``Recent work on understanding how PBT is used in practice has determined that practitioners often run PBT with very short time budgets...''} Given this practical constraint, it is imperative that 
%inputs be generated as fast as possible; in other words, generators must be 
%treated as performance-sensitive code.

@@

  % in strict functionallanguages like OCaml and Scal
To address this question, we propose a two-pronged approach for designing
generator libraries that preserves their expressiveness while eliminating
unnecessary overhead.
First, we describe a method of optimizing generator libraries using
\textit{multi-stage programming}, or ``staging,'' which has long been used for
optimizing idiomatic functional DSLs~\cn.
We extend this technique to generators, completely eliminating the overhead of
many common generator abstractions. For example, in unstaged generator
libraries, each call to monadic bind usually spends time constructing and
immediately executing a closure at runtime. \jwc{Todo: joe --- cite some evidence for this}
Meanwhile, a staged approach can resolve the bind at compile time, enabling the
compiler to inline and eliminate these closures.
Secondly, we demonstrate the impact of different choices of \rand on generator
performance by conducting a controlled comparison. \hg{More words here for
parallelism?} Together, these insights provide a recipe for improving even
state-of-the-art generator libraries.

@@

To demonstrate the effectiveness of our technique, we apply it to \bq, a
high-quality PBT library in OCaml, authored by Jane Street.
Our improved version, \name, is a direct drop-in replacement for
\bq that 
is \textit{completely semantically equivalent}: that is, given identical random seeds, generators written using \name{}
produce exactly the same sequence of values as those written in \bq. However, in \name, 
the abstractions provided by the generator DSL are zero-cost, and the \rand is 
highly-optimized. The direct equivalence between \name and \bq allows us to 
perform extremely fine-grained comparisons, differences in generation speed are
directly attributable to the staging and \rand optimizations.

>>>>

Property-based testing (PBT) is a technique for software testing which uses
random test inputs to evaluate logical properties that the system is expected to
satisfy.  Recent studies on PBT usage~\cite{inpractice} have shown that
practitioners interleave PBT with their software development process, sometimes
testing their properties every time they save their code. This workflow makes
sense because the developer who just wrote the code is often the best person to
fix a potental bug; if a bug is found later, it will need to be fixed by someone
with less context. \jwc{Not sure how the previous sentence is relevant. It delays us getting to the point.}  All of this means that PBT is often run with very short time
budgets --- e.g., between 50 milliseconds and 30 seconds --- and it is important
that bugs can be found in that time-frame.
\jwc{phrasing could suggest ``race against time'' aspect.}\hg{Agree, can't see
how to fit this in yet}

@@

We find that the performance of existing generator libraries falls short of what
is possible. This is due to two major sources of inefficiency.
First, the same high-level design that gives generator libraries their
flexibility and expressiveness also introduces overhead: layered abstractions
and indirections hinder the compiler's optimizations, shifting more work to
run time.
Second, the \rand that underlies the generator's operation --- the source of
random numbers that are used to make choices in the generator --- 
can constitute an unexpectedly high proportion of a generator's running time.
\iflater
\hg{TODO later, I'm not 100\% sure it's clear how this particular question is
raised by the above. We could strengthen that connection}
\fi
These observations together raise a fundamental question: can generator
libraries produce code that is both expressive \textit{and} efficient?
%In practice, PBT is used 
%as a lightweight, ad hoc approach to software correctness,
%\jwc{Move this bit about how it's used to the first sentence. I.e. PBT is a testing technqique where random data is fed to a system under test...}
%aimed at 
%detecting bugs within a short testing window---typically 50 milliseconds 
%to 30 seconds. \jwc{This sells us a little short: we should say something to the effect of ``Recent work on understanding how PBT is used in practice has determined that practitioners often run PBT with very short time budgets...''} Given this practical constraint, it is imperative that 
%inputs be generated as fast as possible; in other words, generators must be 
%treated as performance-sensitive code.

@@

  % in strict functionallanguages like OCaml and Scal
To address this question, we propose a two-pronged approach for designing
generator libraries that preserves their expressiveness while eliminating
unnecessary overhead.
First, we describe a method of optimizing generator libraries using
\textit{multi-stage programming}, or ``staging,'' which has long been used for
optimizing idiomatic functional DSLs~\cn.
We extend this technique to generators \jwc{``apply this technology to''? We're not extending staging, we're a client}, completely eliminating the overhead of
many common generator abstractions. For example, in unstaged generator
libraries, each call to monadic bind usually spends time constructing and
immediately executing a closure at runtime. \jwc{Todo: joe --- cite some evidence for this}
Meanwhile, a staged approach can resolve the bind at compile time, enabling the
compiler to inline and eliminate these closures.
Secondly, we demonstrate the impact of different choices of \rand on generator
performance by conducting a controlled comparison. \hg{More words here for
parallelism?} Together, these insights provide a recipe for improving even
state-of-the-art generator libraries.

@@

To demonstrate the effectiveness of our technique, we apply it to \bq, a
high-quality PBT library in OCaml, authored by Jane Street.
Our improved version, \name, is a direct drop-in replacement for
\bq that 
is \textit{completely semantically equivalent}: that is, given identical random seeds, generators written using \name{}
produce exactly the same sequence of values as those written in \bq. However, in \name, 
the abstractions provided by the generator DSL are zero-cost, and the \rand is 
highly-optimized. The direct equivalence between \name and \bq allows us to 
perform extremely fine-grained comparisons, \propose{as} differences in \proposechange{generation speed}{bugfinding power} are
\proposechange{directly}{solely} attributable to the staging and \rand optimizations.
